---
title:  "Preparing a WHAM input"
subtitle: ""
author: "Tim Miller<br>NOAA Fisheries, NEFSC"
output:
  xaringan::moon_reader:
    self_contained: true
    css: ["xaringan-themer_16_9.css", "slides-style_TMB201.css"]
    lib_dir: libs
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

layout: true

.footnote[U.S. Department of Commerce | National Oceanic and Atmospheric Administration | National Marine Fisheries Service]


<style type="text/css">

code.cpp{
  font-size: 14px;
}
code.r{
  font-size: 14px;
}


</style>

```{css, echo=FALSE}
pre {
  max-height: 250px; /*changes height of chunk output box*/
  max-width: 800px; /*changes width of chunk output box*/
  overflow-y: auto; /* auto will add vertical scroll bar when necessary */
}

```


```{r setup, include=FALSE}
#options(htmltools.dir.version = FALSE)
#library(knitr)
#knitr::opts_chunk$set(cache = TRUE)
#knitr::opts_knit$set(root.dir = here::here())
#knitr::opts_chunk$set(width = 1500)
#library(kableExtra)
#library(imager)
#here::i_am("c:/work/BSB_RT")

#library(here)

```

```{r set-options, include = FALSE}
options(width = 50)
```

```{r xaringan-tile-view, echo=FALSE}
# this gives you a tile navigation if you type "O" at any time
#xaringanExtra::use_tile_view()
```

---
# Outline <br>

Creating an input for fitting a model with WHAM:
* From data files
* Modifying an existing WHAM input
* Using ASAP input
---

# Bulding an ASAP Input

* First, we need to read in all of the necessary data inputs for WHAM. 
* These inputs were created from the ASAP input `ex2_SENMAYT.dat` file.
* We will see that the results are the same whether we use the ASAP dat file directly or cobble the input together by hand.

---

# Bulding an ASAP Input

First the biological information:

```{r, include = FALSE}
library(here)
```
```{r, eval = FALSE}
library(here)
```
```{r, eval = TRUE}
n_regions <- 1L
n_stocks <- 1L

#just one maturity at age matrix (1 stock)
temp <- lapply(1:n_stocks, \(i) as.matrix(read.csv(here("data", paste0("mat_",i,".csv")))))
mat <- array(0, dim = c(n_stocks,dim(temp[[1]])))
for(i in 1:n_stocks) mat[i,,] <- temp[[i]]

dim(mat) #(n_stocks x n_years x n_ages)
n_years <- dim(mat)[2]
n_ages <- dim(mat)[3]

years = 1973:2011 

ages <- 1:n_ages
```

---

# Bulding an ASAP Input

More biological information:

```{r, eval = TRUE}
#there are 7 WAA matrices
temp <- lapply(1:7, \(i) as.matrix(read.csv(here("data", paste0("waa_",i,".csv")))))
waa <- array(0, dim = c(7,dim(temp[[1]])))
for(i in 1:(dim(waa)[1])) waa[i,,] <- temp[[i]]

waa_pointer_ssb <- 7

#1 column for 1 stock
fracyr_ssb <- as.matrix(read.csv(here("data", "fracyr_ssb.csv")))

#Make MAA array (n_stocks x n_regions x n_years x n_ages)
MAA <- array(NA, dim = c(n_stocks, n_regions, n_years, n_ages))
for(i in 1:n_stocks) for(j in 1:n_regions) MAA[i,j,,] <- as.matrix(read.csv(file = here("data", paste0("MAA_stock_", i, "_region_",j,".csv"))))
```
---

# Bulding an ASAP Input

Next the catch information:
```{r, eval = TRUE}
catch <- as.matrix(read.csv(here("data", "catch.csv")))
any(!catch>0) #cannot be any years without catch

#should be n_years x n_fleets
dim(catch) 

#just one column: 1 fleet
n_fleets <- NCOL(catch)

catch_cv <- as.matrix(read.csv(here("data", "catch_cv.csv")))
any(!catch_cv>0) #cannot be any years with cv missing

#should be n_years x n_fleets
dim(catch_cv)
```

---

# Bulding an ASAP Input

More catch infomation:
```{r, eval = TRUE}
#proportions at age matrix for each fleet is n_years x n_ages
catch_paa <- as.matrix(read.csv(here("data", "catch_paa_fleet_1.csv")))
dim(catch_paa) 

#array used by WHAM: (n_fleets x n_years x n_ages)
catch_paa <- array(catch_paa, dim = c(1,dim(catch_paa)))
catch_paa[which(catch_paa<0)] <- NA

catch_Neff <- as.matrix(read.csv(here("data", "catch_Neff.csv")))

#should be n_years x n_fleets
dim(catch_Neff)

#0/1 matrx (n_years x n_fleets) whether to use proportions at age each year)
use_catch_paa <- matrix(0L, n_years, n_fleets)
for(f in 1:n_fleets) for(y in 1:n_years) use_catch_paa[y,f] <- sum(!any(is.na(catch_paa[f,y,])))

#Neff must be greater than 0)
use_catch_paa[which(!catch_Neff>0)] <- 0

#same selectivity model for the fleet for all years
selblock_pointer_fleets <- matrix(1, n_years, n_fleets)

#first WAA matrix
waa_pointer_fleets <- 1
```

---

# Bulding an ASAP Input

Finally, the information for all indices:
```{r, eval = TRUE}
indices <- as.matrix(read.csv(here("data", "indices.csv")))
index_cv <- as.matrix(read.csv(here("data", "index_cv.csv")))
#indices must be greater than 0
indices[which(!indices>0)] <- NA
indices[which(!index_cv>0)] <- NA

#should be n_years x n_indices
dim(indices)

#5 indices
n_indices <- NCOL(indices)

#How is the index measured? 2 = numbers, 1 = biomass
units_indices <- c(2,2,2,1,1)
#How are the proportions at age measured? 2 = numbers, 1 = biomass
units_index_paa <- c(2,2,2,1,1)

index_Neff <- as.matrix(read.csv(here("data", "index_Neff.csv")))

#should be n_years x n_indices
dim(index_Neff)
```

---

# Bulding an ASAP Input

More index information:

```{r, eval = TRUE}

#different selectivity model for each index
selblock_pointer_indices <- t(matrix(1+1:n_indices, n_indices, n_years))

#proportions at age matrix for each index is n_years x n_ages
temp <- lapply(1:n_indices, \(i) as.matrix(read.csv(here("data", paste0("index_paa_",i,".csv")))))

#array used by WHAM: (n_indices x n_years x n_ages)
index_paa <- array(0, dim = c(n_indices,dim(temp[[1]])))
for(i in 1:n_indices) index_paa[i,,] <- temp[[i]]

#0/1 matrix (n_years x n_fleets) whether to use indices and proportions at age each year)
use_index_paa <- use_indices <- matrix(0L, n_years, n_indices)

use_indices[] <- as.integer(!is.na(indices))
for(i in 1:n_indices) for(y in 1:n_years) use_index_paa[y,i] <- sum(!any(is.na(index_paa[i,y,])))
#Neff must be greater than 0)
use_index_paa[which(!index_Neff>0)] <- 0


index_fracyr <- as.matrix(read.csv(here("data", "index_fracyr.csv")))

#should be n_years x n_indices
dim(index_fracyr)

waa_pointer_indices <- 1  + 1:n_indices
```

---


# Bulding an ASAP Input

Now we are ready to build the necessary arguments required by `prepare_wham_input`. 

First is the `basic_info` argument which includes biological information. Note, we now provide information on the number of stocks, regions, and intervals, but default values are the same.
```{r, eval = TRUE}
basic_info <- list(
  n_stocks = n_stocks,
  ages = ages,
  n_fleets = n_fleets,
  fracyr_SSB = fracyr_ssb,
  maturity = mat,
  years = years, 
  waa = waa,
  waa_pointer_ssb = waa_pointer_ssb
)

```

---

# Bulding an ASAP Input

Second, the catch information:
```{r, eval = TRUE}
catch_info <- list(
  n_fleets = n_fleets,
  agg_catch = catch,
  agg_catch_cv = catch_cv,
  catch_paa = catch_paa,
  use_catch_paa = use_catch_paa,
  catch_Neff = catch_Neff,
  selblock_pointer_fleets = selblock_pointer_fleets,
  waa_pointer_fleets = waa_pointer_fleets
)

```
---

# Bulding an ASAP Input

Next, the information for the indices:
```{r, eval = TRUE}
index_info <- list(
  n_indices = n_indices,
  agg_indices = indices,
  units_indices = units_indices,
  units_index_paa = units_index_paa,
  agg_index_cv = index_cv,
  fracyr_indices = index_fracyr,
  use_indices = use_indices,
  use_index_paa = use_index_paa,
  index_paa = index_paa,
  index_Neff = index_Neff,
  selblock_pointer_indices = selblock_pointer_indices,
  waa_pointer_indices = waa_pointer_indices
)
```
---

# Bulding an ASAP Input

The last two arguments to prepare_wham_input that we need are for natural mortality and selectivity. Natural mortality is separated from basic biology because there are many different ways to model natural mortality in WHAM.

Natural mortality is not estimated. Assumed M changes with age for this stock, but is constant over time.
```{r, eval = TRUE}
M_in <- list(initial_MAA = MAA)
```

Specify using logistic selectivity for all 6 selectivity blocks and initial values for $a_{50}$ and inverse of slope. Fixing selectivity for the last two selectivity blocks.
```{r, eval = TRUE}
selectivity <- list(
  model = rep("logistic",6), n_selblocks = 6,
  fix_pars = list(NULL,NULL,NULL,NULL, 1:2, 1:2),
  initial_pars = list(c(2,0.2),c(2,0.2),c(2,0.2),c(2,0.2),c(1.5,0.1),c(1.5,0.1))
)
```

---

# Creating the WHAM input

First, load the WHAM package. 
This will find and load the wham package that is first in the search list of locations that R checks.
```{r, eval = TRUE}
library(wham)
```

We can build all at once using the arguments in `prepare_wham_input`:
```{r, eval = FALSE}
input_all <- prepare_wham_input(basic_info = basic_info, selectivity = selectivity, catch_info = catch_info, index_info = index_info, M = M_in)
```

Build input sequentially using `set_*` functions:
```{r, eval = FALSE}
input_seq <- prepare_wham_input(basic_info = basic_info)
input_seq <- set_M(input_seq, M = M_in)
input_seq <- set_catch(input_seq, catch_info = catch_info)
input_seq <- set_F(input_seq)
input_seq <- set_indices(input_seq, index_info = index_info)
input_seq <- set_q(input_seq)
input_seq <- set_selectivity(input_seq, selectivity = selectivity)
```
```{r, include = FALSE}
input_all <- prepare_wham_input(basic_info = basic_info, selectivity = selectivity, catch_info = catch_info, index_info = index_info, M = M_in)
input_seq <- prepare_wham_input(basic_info = basic_info)
input_seq <- set_M(input_seq, M = M_in)
input_seq <- set_catch(input_seq, catch_info = catch_info)
input_seq <- set_F(input_seq)
input_seq <- set_indices(input_seq, index_info = index_info)
input_seq <- set_q(input_seq)
input_seq <- set_selectivity(input_seq, selectivity = selectivity)
```

---
# Using ASAP Input <br>

A mentioned , we used data from an existing ASAP dat file that is provided in the WHAM package.

```{r, eval = FALSE}
path_to_examples <- system.file("extdata", package="wham")
asap3 <- read_asap3_dat(file.path(path_to_examples,"ex2_SNEMAYT.dat"))
input_asap <- prepare_wham_input(asap3) 
```
```{r, include = FALSE}
path_to_examples <- system.file("extdata", package="wham")
asap3 <- read_asap3_dat(file.path(path_to_examples,"ex2_SNEMAYT.dat"))
input_asap <- prepare_wham_input(asap3) 
```

---

# Basic WHAM model

With no other arguments specified to the function:
* a basic statistical catch-at-age model will be configured. 
 * No random effects
 * Annual recruitment treated as fixed effects.
 * Stock-recruit function is not allowed.
* When using ASAP dat files, WHAM defaults match the ASAP3 configuration as closely as possible.
 * Same data, CV, effective sample size assumptions.
 * Typically same assumptions for initial parameter values.
 * BUT any ASAP3 penalties and phases are ignored.


---

# Inside the WHAM input

The object returned by `prepare_wham_input` is a list of objects. 
```{r, eval = TRUE}
names(input_asap)
```
* Some elements are used by `TMB::MakeADFun`: data, par, map, random
* Others provide information for generating output: years, years_full, ages.lab, index_names, fleet_names, etc. 
* `log` is the list of text strings printed to the screen by `prepare_wham_input`
* `options` are the lists provided for the various arguments to `prepare_wham_input` or the various `set_*` functions.
* `call` saves the call expression to `prepare_wham_input`

---

# Compare unfit models


```{r, eval = FALSE}
nofit_all <- fit_wham(input_all, do.fit = FALSE)
nofit_seq <- fit_wham(input_seq, do.fit = FALSE)
nofit_asap <- fit_wham(input_asap, do.fit = FALSE)
```

```{r, include = FALSE}
nofit_all <- fit_wham(input_all, do.fit = FALSE)
nofit_seq <- fit_wham(input_seq, do.fit = FALSE)
nofit_asap <- fit_wham(input_asap, do.fit = FALSE)
```

We can see that the negative log likelihood is the same whether we build the input all at once or sequentially, but not directly from the asap dat file.
```{r, eval = TRUE}
nofit_all$fn() - nofit_seq$fn() #equal
nofit_all$fn() - nofit_asap$fn() #different
```
---

# Compare unfit models

The asap file includes different initial values for initial abundance at age and catchability:
```{r, eval = TRUE}
nofit_asap$par-nofit_all$par
```
Evaluating from the same initial parameters give the same nll:
```{r, eval = TRUE}
nofit_asap$fn(nofit_all$par)-nofit_all$fn()
```

---

# Compare fitted models

And you can verify that the optimized models using each of the 3 inputs gives the same results.

```{r, eval = FALSE}
fit_seq <- fit_wham(input_seq, do.retro = FALSE, do.osa = FALSE, do.sdrep = FALSE)
fit_all <- fit_wham(input_all, do.retro = FALSE, do.osa = FALSE, do.sdrep = FALSE)
fit_asap <- fit_wham(input_asap, do.retro = FALSE, do.osa = FALSE, do.sdrep = FALSE)
```
```{r, include = FALSE}
res_dir <- file.path("../temp")
fit_seq <- readRDS(file.path(res_dir,"fit_seq.RDS"))
fit_all <- readRDS(file.path(res_dir,"fit_all.RDS"))
fit_asap <- readRDS(file.path(res_dir,"fit_asap.RDS"))
```

```{r, eval = TRUE}
fit_seq$opt$obj - fit_all$opt$obj # equal
fit_asap$opt$obj - fit_all$opt$obj # equal
```

